Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-19 paper.tex
minors

\usepackage[T1]{fontenc}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage[ruled]{algorithm2e}
\def\doi#1{\href{https://doi.org/\detokenize{#1}}{\url{https://doi.org/\detokenize{#1}}}}
%
\usepackage{graphicx}

@@

The use of \acp{GNN} as part of a distributed controller has been explored in previous literature~\cite{DBLP:conf/icra/GosrichMLPYR022,DBLP:conf/corl/TolstayaGPP0R19}, 
 where it was shown that they could be used to break down the evaluation of local and distributed systems. 
However, in these works, communication was left entirely to the neural network, 
 making the learning process potentially more complex and unstable. 
 In our approach, the \ac{GNN} is \emph{informed} by computational fields that collect the necessary information to compute a certain task, 
 limiting learning to only the specific task defined by a collective reward function.
 This will speed up the learning process and make it more stable.
\section{Field-informed Reinforcement Learning}
\label{sec:approach}
%\ga{Page budget: 2/3 pages \\}
%\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics. Particularly, I would like to highlight how each node in the graph is a local controller, but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}
\begin{figure}
	\centering
  \includegraphics[width=.98\linewidth]{imgs/architecture.pdf}
  \label{fig:architecture}

@@

  For each time step $t$, a graph is constructed from the environment, and it will associate each node with a local feature (hexagons in the picture). 
  Using this feature, an aggregate program computes spatio-temporal information that enhances the feature set of each agent, depicted as colours in the middle graph. 

@@

@@

The proposed solution consists mainly of two parts, 

@@

  \caption{Simulations of the case study scenario in Alchemist. 
  The dots represent the agent, 
  the circle area on represents the phenomenon to be monitored.
  \Cref{fig:static} represents the scenario used during training. 
  The others are used in the test phase to evaluate the policy found. }
  \label{fig:scenarios}
\end{figure}
%\ga{Page budget: 2 pages \\}
%\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
%I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.}
To test the effectiveness of the proposed approach, 
 we decided to experiment with a case study related to swarm robotics, 
 specifically the tracking and coverage of a spatio-temporal phenomenon, cf. tracking a wildfire 
 or monitoring the water levels in a canal with multiple autonomous agents embodied in embedded devices (e.g., drones or IoT devices). The individual agents do not have any knowledge of the initial phenomenon itself (i.e., shape, size, location, velocity,...).
 %
Initially, 
  we perform the training phase using a stationary phenomenon before expanding towards a moving phenomenon in the test phase. 
  %Additionally, the phenomenon can change its shape during runtime. 
  Finally, the phenomenon may have varying areas of interest, defined by an underlying distribution function. This underlying distribution is utilised in the feature set of each agent's observation and guides the agents to rally over the phenomenon. While we only use Guassian distributions but we can use any other distribution and shape. 
As a simulation environment, we use Alchemist~\cite{alchemist}, 
 a simulator for multi-agent systems that allows us to simulate the swarm behaviour of the agents 
 and the phenomenon of interest.
For the \ac{GNN}, we use the implementation provided by PyTorch Geometric~\cite{Fey/Lenssen/2019}, 
 which is a library for deep learning on graphs built on top of PyTorch~\cite{torch}.
Finally, we use ScaFi~\cite{casadei2022scafi} as the aggregate programming language to support our field-informed approach.
The evaluation is performed in two stages, first the neural networks are trained in an explicit training phase before extensively evaluated in the testing stage.

\subsection{Scenario}
Figure~\ref{fig:scenarios} presents the three different types of scenarios utilised within the evaluation. The first type of experiments considers a single phenomenon at a static location (i.e., \emph{Zone Fixed}), the second type of experiment considers two phenomena in two independent but static locations (i.e., \emph{Two Zones}), and the third type of experiment considers a moving phenomenon (i.e., \emph{Moving}). All phenomena are modelled as a Gaussian distribution. Importantly, only the left scenario illustrated in Figure~\ref{fig:static} was used for training the neural networks. 
Furthermore, all three types of scenarios contain a set of $\mathcal{P} = 25$ agents placed in a 2D grid large 1000x1000 meters. % and a single, stationary phenomenon with a  
Each agent can perceive the presence
 of the phenomenon of interest through an installed sensor $\zeta_v$ with $v \in \mathbb{V}$. 
 (e.g., camera, temperature sensor, etc.) if it is within range.
%
Additionally, each agent has a coverage range $\omega$ (fixed to 75 meters) 
 that describes the area it can monitor. 
 Each agent can only communicate directly with its own neighbourhood $\mathcal{N}$, 
 which in this case depends on a $\mathbb{O}$ range fixed to 300 meters. %\lukas{$\mathbb{O}$  not defined.}
Through this communication channel, agents can exchange information. 
%From the neighbourhood, each agent can also perceive the direction of the other agents.
%
Each agent moves following a certain action composed of two components (r, i) 
 which respectively describe the angle and intensity of the movement (i.e., the velocity vector).
Since we used a value-based approach, 
 the action space $\mathcal{A}$ is discrete and composed of 18 possible angles and 3 possible intensities.
In particular, the angles are quantized to 20 degrees, and for the velocities, we have selected [0, 5, 10] m/s.
%\lukas{is this training the \ac{GNN}? }, 
%

For the aggregated information, each agent will produce a computation field with which they will try to approximate the direction of the phenomenon of interest. 
%

%The specific program in question is a simple application of block G, where the source is the maximum value of the neighbourhood. 
The program $\Gamma$ in question is a simple application of block G, where the source is the maximum value of the neighbourhood. 
This can be expressed in Scafi as follows:
\begin{lstlisting}[mathescape=true]
val source = maxHood(nbr(sense($\zeta$))) == sense($\zeta$)
G(source, Point3D.Zero, _ + nbrVector())
\end{lstlisting}
where \texttt{maxHood} is a function that returns the maximum value of the neighbourhood, 
 \texttt{nbr} is a function that returns the value of the neighbourhood, 
\texttt{sense} is a function that returns the value of the sensor, and
\texttt{nbrRange} is a function that returns an approximate direction for each agent in the neighbourhood.
%
This value will then be fed into the $\pi_{GNN}$ to compute the action to be performed.
 %\lukas{not sure if this paragraph should be its own subsection. or at least it needs  different title.}

@@

The \ac{MLP} uses the same feature set as the \ac{GNN} but applies it in the \ac{DQN} but 
%in the proposed approach, but differently, it uses a standard DQN approach 
without leveraging the graph structure. Moreover, we increase the batch size to 512 and the replay buffer to 10000 since we record 25 agents' experiences for each step instead of one graph experience.
%\lukas{to make sure: in all three cases we use DQN but with (i) MLP, (ii) standard GNN, (iii) our approach - a field-informed GNN, right?}
%
%The \ac{GNN} instead of using the field information, use directly the position of nodes and the local sensor value as input features.
The \ac{GNN} alone, without using the field information, apply the position of agents and the local sensor value directly as input features within the \ac{DQN}.
%
These baselines are used to verify the effectiveness of the components used in the field-informed \ac{RL}. 
%
Indeed the \ac{MLP} baseline is used to verify the effectiveness of the \ac{GNN} in the proposed approach, while the \ac{GNN} baseline is used to verify the effectiveness of the field information in the proposed approach.
\subsection{Metrics}
We evaluate the performance of the different approaches by measuring the coverage of the phenomenon over time.
%
The coverage is defined as the percentage of the phenomenon covered by the agents.
%
Specifically, we can measure the coverage as the intersection over the union of the phenomenon and the agents' view range.
%
Formally, first, we define the overall coverage for a certain time step as:
\begin{equation*}
\Omega = \bigcup_{v \in V} \omega_v
\quad 
C = \frac{|\Omega \cap \mathcal{P}|}{|\mathcal{P}|}
\end{equation*}
where $\mathcal{P}$ is the area of the phenomenon and $\Omega$ is the area covered by the agents.
%This is computed for each step of the simulation. 
In the training phases, 
 we measure the average coverage in each episode, and the total reward obtained by the agents at each episode of the simulation. 
%
We also measure the number of agents that are within the phenomenon at each step of the simulation.
This will be a measure of how well the agents are tracking the phenomenon.
\ga{todo improve the description of the coverage}

>>>>

\usepackage[T1]{fontenc}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{acronym}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage[ruled]{algorithm2e}
\def\doi#1{\href{https://doi.org/\detokenize{#1}}{\url{https://doi.org/\detokenize{#1}}}}
%
\usepackage{graphicx}

@@

The use of \acp{GNN} as part of a distributed controller has been explored in previous literature~\cite{DBLP:conf/icra/GosrichMLPYR022,DBLP:conf/corl/TolstayaGPP0R19}, 
 where it was shown that they could be used to break down the evaluation of local and distributed systems. 
%
However, in these works, communication was left entirely to the neural network, 
 making the learning process potentially more complex and unstable. 
 In our approach, the \ac{GNN} is \emph{informed} by computational fields that collect the necessary information to compute a certain task, 
 limiting learning to only the specific task defined by a collective reward function.
 This will speed up the learning process and make it more stable.
\section{Field-informed Reinforcement Learning}
\label{sec:approach}
%\ga{Page budget: 2/3 pages \\}
%\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics. Particularly, I would like to highlight how each node in the graph is a local controller, but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}
\begin{figure}
	\centering
  \includegraphics[width=.98\linewidth]{imgs/architecture.pdf}

@@

  For each time step $t$, a graph is constructed from the environment, and it will associate each node with a local feature $o^v_t$ (hexagons in the picture). 
  Using this feature, an aggregate program computes spatio-temporal information that enhances the feature set of each agent producing $f_v$, depicted as colours in the middle graph. 

@@

}  \label{fig:architecture}

@@

The proposed solution, summarised in \Cref{fig:architecture}, consists mainly of two parts, 

@@

  \caption{Simulations of the case study scenario in Alchemist. 
  The dots represent the agent, 
  the circle area on represents the phenomenon to be monitored.
  \Cref{fig:static} represents the scenario used during training. 
  The others are used in the test phase to evaluate the policy found. }
  \label{fig:scenarios}
%\ga{Page budget: 2 pages \\}
%\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
%I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.}
To test the effectiveness of the proposed approach, 
 we decided to experiment with a case study related to swarm robotics, 
 specifically the tracking and coverage of a spatio-temporal phenomenon, cf. tracking a wildfire 
 or monitoring the water levels in a canal with multiple autonomous agents embodied in embedded devices (e.g., drones or IoT devices). The individual agents do not have any knowledge of the initial phenomenon itself (i.e., shape, size, location, velocity,...).
Initially, 
  we perform the training phase using a stationary phenomenon before expanding towards a moving phenomenon in the test phase. 
  %Additionally, the phenomenon can change its shape during runtime. 
  Finally, the phenomenon may have varying areas of interest, defined by an underlying distribution function. This underlying distribution is utilised in the feature set of each agent's observation and guides the agents to rally over the phenomenon. While we only use Guassian distributions but we can use any other distribution and shape. 
As a simulation environment, we use Alchemist~\cite{alchemist}, 
 a simulator for multi-agent systems that allows us to simulate the swarm behaviour of the agents 
 and the phenomenon of interest.
For the \ac{GNN}, we use the implementation provided by PyTorch Geometric~\cite{Fey/Lenssen/2019}, 
 which is a library for deep learning on graphs built on top of PyTorch~\cite{torch}.
Finally, we use ScaFi~\cite{casadei2022scafi} as the aggregate programming language to support our field-informed approach.
The evaluation is performed in two stages, first the neural networks are trained in an explicit training phase before being extensively evaluated in the testing stage.
The simulations are publicly available at \url{https://github.com/AggregateComputing/experiment-2023-acsos-field-informed-rl}.
\subsection{Scenario}
\Cref{fig:scenarios} presents the three different types of scenarios utilised within the evaluation. The first type of experiments considers a single phenomenon at a static location (i.e., \emph{Zone Fixed}), the second type of experiment considers two phenomena in two independent but static locations (i.e., \emph{Two Zones}), and the third type of experiment considers a moving phenomenon (i.e., \emph{Moving}). All phenomena are modelled as a Gaussian distribution. Importantly, only the left scenario illustrated in Figure~\ref{fig:static} was used for training the neural networks. 
Furthermore, all three types of scenarios contain a set of $\mathcal{P} = 25$ agents placed in a 2D grid large 1000x1000 meters. % and a single, stationary phenomenon with a  
Each agent can perceive the presence
 of the phenomenon of interest through an installed sensor $\zeta_v$ with $v \in \mathbb{V}$. 
 (e.g., camera, temperature sensor, etc.) if it is within range.
Additionally, each agent has a coverage range $\omega$ (fixed to 75 meters) 
 that describes the area it can monitor. 
 Each agent can only communicate directly with its own neighbourhood $\mathcal{N}$, 
 which in this case depends on a $\mathbb{O}$ range fixed to 300 meters. %\lukas{$\mathbb{O}$  not defined.}
Through this communication channel, agents can exchange information. 
%From the neighbourhood, each agent can also perceive the direction of the other agents.
Each agent moves following a certain action composed of two components (r, i) 
 which respectively describe the angle and intensity of the movement (i.e., the velocity vector).
Since we used a value-based approach, 
 the action space $\mathcal{A}$ is discrete and composed of 18 possible angles and 3 possible intensities.
In particular, the angles are quantized to 20 degrees, and for the velocities, we have selected [0, 5, 10] m/s.
%\lukas{is this training the \ac{GNN}? }, 

For the aggregated information, each agent will produce a computation field with which they will try to approximate the direction of the phenomenon of interest. 
%
%
%The specific program in question is a simple application of block G, where the source is the maximum value of the neighbourhood. 
The program $\Gamma$ in question is a simple application of block G, where the source is the maximum value of the neighbourhood. 
This can be expressed in Scafi as follows:
\begin{lstlisting}[mathescape=true]
val source = maxHood(nbr(sense($\zeta$))) == sense($\zeta$)
G(source, Point3D.Zero, _ + nbrVector())
\end{lstlisting}
where \texttt{maxHood} is a function that returns the maximum value of the neighbourhood, 
 \texttt{nbr} is a function that returns the value of the neighbourhood, 
\texttt{sense} is a function that returns the value of the sensor, and
\texttt{nbrRange} is a function that returns an approximate direction for each agent in the neighbourhood.
%
This value will then be fed into the $\pi_{GNN}$ to compute the action to be performed.
 %\lukas{not sure if this paragraph should be its own subsection. or at least it needs  different title.}

@@

The \ac{MLP} uses the same feature set as the \ac{GNN} but applies it in the \ac{DQN} but 
%in the proposed approach, but differently, it uses a standard DQN approach 
without leveraging the graph structure. Moreover, we increase the batch size to 512 and the replay buffer to 10000 since we record 25 agents' experiences for each step instead of one graph experience.
%\lukas{to make sure: in all three cases we use DQN but with (i) MLP, (ii) standard GNN, (iii) our approach - a field-informed GNN, right?}
%
%The \ac{GNN} instead of using the field information, use directly the position of nodes and the local sensor value as input features.
The \ac{GNN} alone, without using the field information, apply the position of agents and the local sensor value directly as input features within the \ac{DQN}.
%
These baselines are used to verify the effectiveness of the components used in the field-informed \ac{RL}. 
%
Indeed the \ac{MLP} baseline is used to verify the effectiveness of the \ac{GNN} in the proposed approach, while the \ac{GNN} baseline is used to verify the effectiveness of the field information in the proposed approach.
\subsection{Metrics}
We evaluate the performance of the different approaches by measuring the coverage of the phenomenon over time.
%
The coverage is defined as the percentage of the phenomenon covered by the agents.
%
Specifically, we can measure the coverage as the intersection over the union of the phenomenon and the agents' view range.
%
Formally, first, we define the overall coverage for a certain time step as:
\begin{equation*}
\Omega = \bigcup_{v \in V} \omega_v
\quad 
C = \frac{|\Omega \cap \mathcal{P}|}{|\mathcal{P}|}
\end{equation*}
where $\mathcal{P}$ is the area of the phenomenon and $\Omega$ is the area covered by the agents.
%This is computed for each step of the simulation. 
In the training phases, 
 we measure the average coverage in each episode, and the total reward obtained by the agents at each episode of the simulation. 
%
We also measure the number of agents that are within the phenomenon at each step of the simulation.
This will be a measure of how well the agents are tracking the phenomenon.
%\ga{todo improve the description of the coverage}
